Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis

### 2018-03-22 paper/finalish/article.tex
get rid of unnecessary constractions 'don't' and 'we've'

While the overall attribution was consistent between the Dirichlet and asymmetric Island models, it would be expected that the conditional probabilities for a given genotype might differ markedly. For those genotypes observed infrequently (or not at all) among the sources the Dirichlet model has little information while the asymmeteric Island model can exploit information from cases with similar (but not identical) genetic profiles. In the case of MLST data with just 7 loci, the majority of human cases and source isolates come from a relatively small number of sequence types which are observed often. Thus the Dirichlet model performs well, as it has sufficient observations to estimate the genotype distribution well where the bulk of the data lie. It is only those genotypes that are rarely observed where it performs poorly, but as they are rarely observed, they don't contribute significantly to the overall attribution. In other circumstances, such as where we have many more than 7 loci, we would expect to have many more rare genotypes, so that the Dirichlet model might provide little useful information. At the extreme example of whole genome MLST (wgMLST) where each isolate would typically be unique, it would provide essentially no information at all. In such circumstances, however, the asymmetric Island model would be expected to still perform well, assuming that information could still be transferred between similar genotypes.

@@

Future research might focus on possible extensions to these models. One direction is to adapt the models with additional covariates, which might include age, occupation, and other risk factors such as contact with animals. For example, there is evidence that children in rural areas are at higher risk of campylobacteriosis through contact with farm animals \cite{MullM, Spencer2012}. Another direction is in expanding the role of water. In these models, we've assumed that water is a source of human campylobacteriosis infection, but water differs from the other food and environmental sources in that it is not an amplifying reservoir for \emph{Campylobacter} \cite{Wagen}. In contrast, genotypes found in water might be expected to originate in the other sources present here, particularly ruminants and wild birds, but also potentially from humans as well via discharge of unprocessed human waste. Hence, water acts as a transmission pathway from sources to humans, being both an end point (reduced water quality from faecal contamination) and a source (human consumption of water, either recreationally or through untreated water supplies). While there is presently little evidence that water is an important source for human campylobacteriosis from the current models, these are fit on sporadic cases of campylobacteriosis. However, water is known as a key source of outbreaks of campylobacteriosis, such as the large outbreak in Havelock North, New Zealand in 2016 where an estimated 5,500 out of 14,000 residents became ill \cite{dia_campy_report1}. Thus, characterising the source of \emph{Campylobacter} found in water has important implications for both water quality and public health.

>>>>

While the overall attribution was consistent between the Dirichlet and asymmetric Island models, it would be expected that the conditional probabilities for a given genotype might differ markedly. For those genotypes observed infrequently (or not at all) among the sources the Dirichlet model has little information while the asymmeteric Island model can exploit information from cases with similar (but not identical) genetic profiles. In the case of MLST data with just 7 loci, the majority of human cases and source isolates come from a relatively small number of sequence types which are observed often. Thus the Dirichlet model performs well, as it has sufficient observations to estimate the genotype distribution well where the bulk of the data lie. It is only those genotypes that are rarely observed where it performs poorly, but as they are rarely observed, they do not contribute significantly to the overall attribution. In other circumstances, such as where we have many more than 7 loci, we would expect to have many more rare genotypes, so that the Dirichlet model might provide little useful information. At the extreme example of whole genome MLST (wgMLST) where each isolate would typically be unique, it would provide essentially no information at all. In such circumstances, however, the asymmetric Island model would be expected to still perform well, assuming that information could still be transferred between similar genotypes.

@@

Future research might focus on possible extensions to these models. One direction is to adapt the models with additional covariates, which might include age, occupation, and other risk factors such as contact with animals. For example, there is evidence that children in rural areas are at higher risk of campylobacteriosis through contact with farm animals \cite{MullM, Spencer2012}. Another direction is in expanding the role of water. In these models, we have assumed that water is a source of human campylobacteriosis infection, but water differs from the other food and environmental sources in that it is not an amplifying reservoir for \emph{Campylobacter} \cite{Wagen}. In contrast, genotypes found in water might be expected to originate in the other sources present here, particularly ruminants and wild birds, but also potentially from humans as well via discharge of unprocessed human waste. Hence, water acts as a transmission pathway from sources to humans, being both an end point (reduced water quality from faecal contamination) and a source (human consumption of water, either recreationally or through untreated water supplies). While there is presently little evidence that water is an important source for human campylobacteriosis from the current models, these are fit on sporadic cases of campylobacteriosis. However, water is known as a key source of outbreaks of campylobacteriosis, such as the large outbreak in Havelock North, New Zealand in 2016 where an estimated 5,500 out of 14,000 residents became ill \cite{dia_campy_report1}. Thus, characterising the source of \emph{Campylobacter} found in water has important implications for both water quality and public health.
